Computing the bicharacter of huge subspaces of polynomials

### 2019-06-11 article/article.tex
Complete section row symmetries

	\section{Introduction}
	This package was developed at the occasion of a collaboration with François Bergeron to study the character of the space $\harmonics_{k,n}$ of diagonal harmonic polynomials on $k$ sets of $n$ variables, as a graded bi-$\GLr \times \Sn$-module. The main achievement is a calculation of this character for $n=6$ \pauline{+ some cases for $n=7$ ?} and $r=5$ (and therefore any $r$) in 45 minutes (and ~15Gb), which led to conjectures on its general value.

@@

	Consider the polynomial ring $R=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables: \pauline{small $x$ for variables, capital $X$ for sets of variables? }
	$$(x_{i,j})_{0\leq i<k \atop 0\leq j<n}$$
	and we denote by $\bm{x}_i = (x_{i1}, x_{i2}, \dots, x_{in})$ the i-th set of variables. We use the convention that bold letters stand for sets of variables and light letters stand for the variables themselves.

@@

	Like before, we consider the smallest subspace containing $V_{\lambda}$ and closed by partial derivation and polarization, except that this time the inert variables do not intervene in the operators. We denote this space by $\mathcal{E}_{\lambda}$. The diagonal harmonics $\harmonics_{n,r}$ are thus a special case of $\mathcal{E}_{\lambda}$ for $\lambda = n$. \\

@@

	\begin{example} Compute $\harmonics_3$ with one row of variables $\bm{x} = (x_1, x_2, x_3)$.
		\begin{sagesilent}
			load("../subspace.py")
		\end{sagesilent}
		\begin{sagecommandline}
			sage: R = QQ['x1,x2,x3']
			sage: M = matrix.vandermonde(R.gens(), R)
			sage: M
			sage: derivatives = [attrcall('derivative', x) for x in R.gens()]
			sage: H_3 = Subspace([M.determinant()], derivatives)
			sage: H_3.basis()
			sage: H_3.dimension()
		\end{sagecommandline}	
	\end{example}

@@

	Once a basis of $\harmonics_{n,r}$ or $\mathcal{E}_\lambda$ is constructed, there still remains to recover its bicharacter. For $\GLr$ that's given by the multigrading. For $\Sn$ some additional work has to be done (e.g. computing the traces of the action of permutations of all cycle types).

@@

	\begin{example} Compute $\mathcal{E}_{2,1}$ with two rows of variables using the multidegree. 
		\begin{sagesilent}
			load("../compute_character.py")
		\end{sagesilent}
		First, we compute $\mathcal{E}_{2,1}$ for one row of variables. 
		\begin{sagecommandline}
			sage: v = vandermonde(Partition([2,1]))
			sage: v
			sage: derivatives = partial_derivatives(v.parent())
			sage: derivatives
			sage: E_21 = Subspace({v.multidegree(): [v]}, derivatives)
			sage: E_21.basis()
			sage: E_21.dimension()
		\end{sagecommandline}
		Then, we use polarization operators to introduce a second set of variables, still with multidegree. 
		\begin{sagecommandline}
			sage: polarizators = polarization_operators(2)
			sage: polarizators
			sage: E_21 = PolarizedSpace(E_21, polarizators)
			sage: E_21.basis()
			sage: E_21.dimension()
			sage: E_21.dimensions()
		\end{sagecommandline}	
	\end{example}

@@

	The action of $\GLr$ commutes with the action of $\Sn$. Therefore, we can compute independently each $\Sn$ isotypic component of $\harmonics_{n,r}$ by starting from a basis of harmonic polynomials that itself is adapted to its decomposition into isotypic components:
	$$\harmonics_{n}=:\bigoplus_\lambda \harmonics^\lambda\,.$$

	In practice, since the polarization operators commute with the Young symmetrizer, we first compute $\harmonics_{n}$ or $\mathcal{E}_\lambda$ for one set of variables and decompose it into isotypic components using the Young symmetrizer. Then we use the polarization operators to obtain the isotypic components of $\harmonics_{n,r}$ (resp. $\mathcal{E}_\lambda$ in $r>1$ sets of variables) from those of $\harmonics_{n}$ (resp. $\mathcal{E}_\lambda$ in 1 set of variables) .

	\begin{example}
		Compute $\harmonics_{3,2} = \mathcal{E}_3$ using the Vandermonde determinant associated to $\lambda=3$. We still use the multigrading. 
		\begin{sagecommandline}
			sage: v = vandermonde(Partition([3]))
			sage: v
			sage: derivatives = partial_derivatives(v.parent())
			sage: E_3 = Subspace({v.multidegree(): [v]}, derivatives)
		\end{sagecommandline}
		At this stage, we project onto the isotypic components of $\Sym{3}$.
		\begin{sagecommandline}
			sage: E_3 = IsotypicComponent(E_3, 3)
			sage: E_3
		\end{sagecommandline}
		The result is now a dictionnary of subspaces.
		\begin{sagecommandline}
			sage: E_3[Partition([3])].basis()
			sage: E_3[Partition([2,1])].basis()
			sage: E_3[Partition([1,1,1])].basis()
		\end{sagecommandline}
		Now we polarize each isotypic component. Note that, \texttt{PolarizedSpace} can receive a dictionary of subspaces, it then consider each subspace separately. 
		\begin{sagecommandline}
			sage: polarizators = polarization_operators(2)
			sage: E_3 = PolarizedSpace(E_3, polarizators)
			sage: E_3[Partition([2,1])].basis()
		\end{sagecommandline}
	We chose to print only the polarized isotypic component of type $21$ to avoid unreadable outputs in the example. 	
	\end{example}

@@

	One thing that can slow down the computation is that the polynomials we are working with tend to be big. For example, in $6$ variables, the Vandermonde determinant has $6!=720$ monomials. This means that we need a lot of space to store them and it takes time when we want to apply operators on them.

	However, for all $p \in \mathcal{E}_\lambda$, p is anti-symmetric with respect to the column stabilizer $Q_\mu$, with $\mu$ the type of the isotypic component of $p$. Indeed, when projecting $p$ on its isotypic component using the Young idempotent $c_\mu$, by definition $a_\mu$ makes $p$ symmetric with respect to the row preserving permutations of $\mu$ and then $b_\mu$ makes $p.a_\mu$ anti-symmetric with respect to the column preserving permutations of $\mu$. 
	It's therefore sufficient to keep one representative monomial per $S_P$ orbit \pauline{how do we guarantee one monomial is enough?} (and do some appropriate straightening whenever a new monomial is created). We chose to keep the smallest representative monomial in lexicographic order. \\

@@

	\begin{example}
		$\Delta_{21} = -x_{2} \theta_{1} + x_{3} \theta_{1} + x_{1} \theta_{2} - x_{3} \theta_{2} - x_{1} \theta_{3} + x_{2} \theta_{3}$ is type $111$.
		Reduce form : $x_1\theta_2$. \\
		$\partial_{x_2}\Delta_{21} = -\theta_1 + \theta_3$ is type $21$. 
		Reduce form : $-\theta_1$ 
	\end{example}

@@

	$H_k^P$ is symmetric w.r.t. permutation of rows, and such permutations
	exchange multigraded components. For example, for $r=3$, the homogeneous components of respective multidegree $4,2,1$ and $2,1,4$ coincide
	up to cycling of the rows. It is therefore sufficient, in principle, to compute homogeneous components with decreasing multidegree.
	To achieve this, polarization operators can be tweaked: whenever the result of a polarization is not of decreasing multidegree, straighten it by applying an appropriate permutation of the rows. It can then easily be shown that those tweaked polarization operator properly generate the whole $\harmonics_k^P$ space from $\harmonics^P$.

	\begin{example}
		$x_1^2 - 2x_1x_2 + 2x_2x_3 - x_3^2 \in \mathcal{H}_{3}$ so $y_1^2 - 2y_1y_2 + 2y_2y_3 - y_3^2 \in \mathcal{H}_{3}$

		$-2x_{1} y_{1} y_{2} + x_{1} y_{2}^{2} \in \mathcal{H}_{3}$ so $-2y_{1} x_{1} x_{2} + y_{1} x_{2}^{2} \in \mathcal{H}_{3}$
	\end{example}

	\textbf{Gain:} an order of magnitude; below are the total dimension of each $\harmonics_5^\lambda$ and its restriction to homogeneous components of decreasing multidegree.

@@

	\section{Applications}
	Ask Francois ...

	\begin{definition} \pauline{find the right place for Steenrod op}
		The \emph{Steenrod operator} $St_{x_a}^k$ for $k>1$ is given by
		$$St_{x_a}^k = \sum_{i=1}^{n} x_{ai} \partial_{x_{ai}}^k$$
		where $x_a$ is a set of variables of $X$. 
	\end{definition}

>>>>

	\section{Introduction}
	This package was developed at the occasion of a collaboration with François Bergeron to study the character of the space $\harmonics_{k,n}$ of diagonal harmonic polynomials on $k$ sets of $n$ variables, as a graded bi-$\GLr \times \Sn$-module. The main achievement is a calculation of this character for $n=6$ \pauline{+ some cases for $n=7$} and $r=5$ (and therefore any $r$) in 45 minutes (and ~15Gb), which led to conjectures on its general value.

@@

	Consider the polynomial ring $R=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables:
	$$(x_{i,j})_{0\leq i<k \atop 0\leq j<n}$$
	and we denote by $\bm{x}_i = (x_{i1}, x_{i2}, \dots, x_{in})$ the i-th set of variables. We use the convention that bold letters stand for sets of variables and light letters stand for the variables themselves.

@@

	Like before, we consider the smallest subspace containing $V_{\lambda}$ and closed by partial derivation and polarization, except that this time the inert variables do not intervene in the operators. We denote this space by $\mathcal{E}_{\lambda}^{(r)}$. The diagonal harmonics $\harmonics_{n,r}$ are thus a special case of $\mathcal{E}_{\lambda}^{(r)}$ for $\lambda = n$. \\

	The character of $\mathcal{E}_{\lambda}^{(r)}$ stabilizes for $r \geq n$ (see \cite{Bergeron2013} or \cite{Bergeron2009}). Therefore, when the precision is not necessary, we will drop the $(r)$ and simply write $\mathcal{E}_{\lambda}$ for $\mathcal{E}_{\lambda}^{(r)}$ with $r \geq n$. \\

@@

	\begin{example} \label{H_3} Compute $\harmonics_3$ with one row of variables $\bm{x} = (x_1, x_2, x_3)$. \pauline{Je n'ai trouvé comment avoir les output en latex avec sauts de ligne automatiques.}
		\begin{sagesilent}
			load("../subspace.py")
		\end{sagesilent}
		\begin{sagecommandline}
			sage: R = QQ['x1,x2,x3']
			sage: M = matrix.vandermonde(R.gens(), R)
			sage: M
			sage: derivatives = [attrcall('derivative', x) for x in R.gens()]
			sage: H_3 = Subspace([M.determinant()], derivatives)
			sage: H_3.basis()
			sage: H_3.dimension()
		\end{sagecommandline}	
	\end{example}

@@

	Once a basis of $\harmonics_{n,r}$ or $\mathcal{E}_\lambda^{(r)}$ is constructed, there still remains to recover its bicharacter. For $\GLr$ that's given by the multigrading. For $\Sn$ some additional work has to be done (e.g. computing the traces of the action of permutations of all cycle types).

@@

	\begin{example} Compute $\mathcal{E}_{2,1}^{(2)}$ using the multidegree. 
		\begin{sagesilent}
			load("../compute_character.py")
		\end{sagesilent}
		First, we compute $\mathcal{E}_{2,1}^{(1)}$, for one row of variables. 
		\begin{sagecommandline}
			sage: v = vandermonde(Partition([2,1]))
			sage: v
			sage: derivatives = partial_derivatives(v.parent())
			sage: derivatives
			sage: E_21 = Subspace({v.multidegree(): [v]}, derivatives)
			sage: E_21.basis()
			sage: E_21.dimension()
		\end{sagecommandline}
		Then, we use polarization operators to introduce a second set of variables, and thus obtain $\mathcal{E}_{2,1}^{(2)}$, still with multidegree. 
		\begin{sagecommandline}
			sage: polarizators = polarization_operators(2)
			sage: polarizators
			sage: E_21 = PolarizedSpace(E_21, polarizators)
			sage: E_21.basis()
			sage: E_21.dimension()
			sage: E_21.dimensions()
		\end{sagecommandline}
		Compared with example \ref{H_3} in which all basis elements are grouped in one list with key $0$, here basis elements are classified by multidegree with the key corresponding. Moreover, we have access to dimension by multidegree. 
	\end{example}

@@

	The action of $\GLr$ commutes with the action of $\Sn$. Therefore, we can compute independently each $\Sn$ isotypic component of $\harmonics_{n,r}$ by starting from a basis of harmonic polynomials that itself is adapted to its decomposition into isotypic components:
	$$\harmonics_{n}=:\bigoplus_\lambda \harmonics_n^\lambda\,.$$

	In practice, since the polarization operators commute with the Young symmetrizer, we first compute $\harmonics_{n}$ or $\mathcal{E}_\lambda^{(1)}$ and decompose it into isotypic components using the Young symmetrizer. Then we use the polarization operators to obtain the isotypic components of $\harmonics_{n,r}$ (resp. $\mathcal{E}_\lambda^{(r)}$, $r>1$) from those of $\harmonics_{n}$ (resp. $\mathcal{E}_\lambda^{(1)}$) .

	\begin{example} \label{multigrading}
		Compute $\harmonics_{3,2} = \mathcal{E}_3^{(2)}$ using the Vandermonde determinant associated to $\lambda=3$. We still use the multigrading. 
		\begin{sagecommandline}
			sage: v = vandermonde(Partition([3]))
			sage: v
			sage: derivatives = partial_derivatives(v.parent())
			sage: E_3 = Subspace({v.multidegree(): [v]}, derivatives)
		\end{sagecommandline}
		At this stage, we project onto the isotypic components of $\Sym{3}$.
		\begin{sagecommandline}
			sage: E_3 = IsotypicComponent(E_3, 3)
			sage: E_3
		\end{sagecommandline}
		The result is now a dictionnary of subspaces.
		\begin{sagecommandline}
			sage: E_3[Partition([3])].basis()
			sage: E_3[Partition([2,1])].basis()
			sage: E_3[Partition([1,1,1])].basis()
		\end{sagecommandline}
		Now we polarize each isotypic component. Note that, \texttt{PolarizedSpace} can receive a dictionary of subspaces, it then consider each subspace separately. 
		\begin{sagecommandline}
			sage: E_3 = PolarizedSpace(E_3, polarization_operators(2))
			sage: E_3[Partition([2,1])].basis()
		\end{sagecommandline}
	We chose to print only the polarized isotypic component of type $21$ to avoid unreadable outputs in the example. 	
	\end{example}

@@

	One thing that can slow down the computation is that the polynomials we are working with tend to be big. For example, for $\lambda \vdash 6$, the Vandermonde determinant $V_\lambda$, has $6!=720$ monomials. This means that we need a lot of space to store them and it takes time when we want to apply operators on them.

	However, for all $p \in \mathcal{E}_\lambda$, p is anti-symmetric with respect to the column stabilizer $Q_\mu$, with $\mu$ the type of the isotypic component of $p$. Indeed, when projecting $p$ on its isotypic component using the Young idempotent $c_\mu$, by definition $a_\mu$ makes $p$ symmetric with respect to the row preserving permutations of $\mu$ and then $b_\mu$ makes $p.a_\mu$ anti-symmetric with respect to the column preserving permutations of $\mu$. 
	It's therefore sufficient to keep one representative monomial per $S_P$ orbit \pauline{$S_P$ orbit?} (and do some appropriate straightening whenever a new monomial is created). We chose to keep the smallest representative monomial in lexicographic order. \\

@@

	\begin{example} ~

		$\Delta_{21} = -x_{2} \theta_{1} + x_{3} \theta_{1} + x_{1} \theta_{2} - x_{3} \theta_{2} - x_{1} \theta_{3} + x_{2} \theta_{3}$ is type $111$.
		Reduced form : $x_1\theta_2$.

		$\partial_{x_2}\Delta_{21} = -\theta_1 + \theta_3$ is type $21$. 
		Reduced form : $-\theta_1$ 
	\end{example}

	In practice, the antisymmetries are exploited first when projecting $\mathcal{E}_\lambda^{(1)}$ onto isotypic components. When applying the Young idempotent, the reduced form of each element is computed. From this step, the polynomials lie in a diagonal polynomial ring with antisymmetries, thus each polynomial carries the information of its antisymmetries.  
	Then, we maintain the elements of the bases in their reduced form during the polarization.

	\begin{example}
			Compute $\harmonics_{3,2} = \mathcal{E}_3^{(2)}$ but this time exploiting antisymmetries. 
		\begin{sagecommandline}
			sage: v = vandermonde(Partition([3]))
			sage: derivatives = partial_derivatives(v.parent())
			sage: E_3 = Subspace({v.multidegree(): [v]}, derivatives)
		\end{sagecommandline}
		At this stage, we project onto the isotypic components of $\Sym{3}$ exploiting antisymmetries. One can check the difference between this computation and the one in example \ref{multigrading}. 
		\begin{sagecommandline}
			sage: E_3 = IsotypicComponent(E_3, 3, use_antisymmetry=True)
			%sage: E_3[Partition([3])].basis()
			sage: E_3[Partition([2,1])].basis()
			sage: E_3[Partition([1,1,1])].basis()
		\end{sagecommandline}
		We can check that the parent of the polynomials carries the information about the antisymmetries. 
		\begin{sagecommandline}
			sage: p = E_3[Partition([2,1])].basis().values()[0][0]
			sage: p
			sage: p.parent()
		\end{sagecommandline}
		Now we polarize each isotypic component still exploiting antisymmetries.  
		\begin{sagecommandline}
			sage: E_3 = PolarizedSpace(E_3, polarization_operators(2))
			sage: E_3[Partition([2,1])].basis()
		\end{sagecommandline}
	\end{example}

@@

	The space $\mathcal{E}_\lambda$ is symmetric with respect to permutation of rows of $X$, and such permutations exchange multigraded components. For example, for $r=3$, the homogeneous components of respective multidegree $(4,2,1)$ and $(2,1,4)$ coincide up to cycling of the rows. This means that there are polynomials that are the same up to a permutation of the sets of variables and that we have redundant information.

	It is therefore sufficient, in principle, to compute homogeneous components with decreasing multidegree. Thus, we reduce memory complexity since not every homogeneous components are kept and time complexity by decreasing the number of polynomials on which apply operators.

	To achieve this, polarization operators can be tweaked: whenever the result of a polarization is not of decreasing multidegree, straighten it by applying an appropriate permutation of the rows. It can then easily be shown that those tweaked polarization operator properly generate the whole $\mathcal{E}_\lambda^{(r)}$ space from $\mathcal{E}_\lambda^ {(1)}$.

	\begin{example} ~

		The polynomial $x_1^2 - 2x_1x_2 + 2x_2x_3 - x_3^2 \in \mathcal{H}_{3}$ has multidegree $(1,0)$ so the same polynomial with multidegree $(0,1)$ that is $y_1^2 - 2y_1y_2 + 2y_2y_3 - y_3^2$ also appears in $\mathcal{H}_{3}$ but it is not necessary to compute it explicitly.  
		%$-2x_{1} y_{1} y_{2} + x_{1} y_{2}^{2} \in \mathcal{H}_{3}$ so $-2y_{1} x_{1} x_{2} + y_{1} x_{2}^{2} \in \mathcal{H}_{3}$
	\end{example}

	In practice, we exploit the row symmetries when we set up the polarization operators. As said above, when specifying the row symmetries, only components of decreasing multidegree appear in the final result.

	\begin{example} Exploiting both antisymmetries of columns and row symmetries, the homogeneous components for $\mathcal{E}_3^{(2)}$ become the following.
		\begin{sagesilent}
			v = vandermonde(Partition([3]))
			derivatives = partial_derivatives(v.parent())
			E_3 = Subspace({v.multidegree(): [v]}, derivatives)
		\end{sagesilent}
		\begin{sagecommandline}
			sage: E_3 = IsotypicComponent(E_3, 3, use_antisymmetry=True)
			sage: E_3 = PolarizedSpace(E_3, polarization_operators(2, row_symmetry="permutation"))
			sage: E_3[Partition([1,1,1])].basis()
			sage: E_3[Partition([2,1])].basis()
			sage: E_3[Partition([3])].basis()
		\end{sagecommandline}	
	We observe that only decreasing multidegree components still appears, we don't have components of multidegree $(1,2)$, $(0,3)$, $(0,2)$ and $(0,1)$.
	\end{example}

	By exploiting this symmetries, we gain an order of magnitude.\\

	As said before, the character of $\GLr$ is given by the multidegree of the homogeneous components. In other words, for a given isotypic component of type $\mu$, the $\GLr$-character associated to this component is given by $\sum_{\mathcal{E}} q^{d(\mathcal{E})}$ where the sum is over all homogeneous $\mathcal{E}$ component of type $\mu$, $d(\mathcal{E})$ is the multidegree of the component $\mathcal{E}$ and $q^d = q_1^{d_1}q_2^{d_2}\dots$.

	When exploiting the row symmetries, since we don't compute only the homogeneous components of decreasing multidegree, we can not compute the character of $\GLr$ using such a summation. However, we know that if we have a component of multidegree $d=(d_1, d_2, \dots, d_k)$ , this correspond to a term $\sum_{\sigma \in \Sym{k}} q^{\sigma(d)} = m_d$ in the previous sum, thus we use monomials symmetric functions instead of summations to recover the character of $\GLr$ in each isotypic component.
	\pauline{Add time computation comparison (with and without optimisations) ?} \\

	For now we don't avoid to compute unnecessary homogeneous component since all we do is straighten the polynomials if they don't have decreasing multidegree. It would be even better if we could avoid computing those polynomials. It is not possible for now because we don't know yet which operators should not been applied to avoid those polynomials.

@@

	\section{Applications}
	Ask Francois.\\
	Talk about non partition diagram, with or without hole (compositions).

	\begin{definition} \pauline{Steenrod operators? }
		The \emph{Steenrod operator} $St_{x_a}^k$ for $k>1$ is given by
		$$St_{x_a}^k = \sum_{i=1}^{n} x_{ai} \partial_{x_{ai}}^k$$
		where $x_a$ is a set of variables of $X$. 
	\end{definition}
